Impact of the High Population Events on Urban Mobility: Who is Affected the Most?

### 2019-12-10 manuscript.Rmd
donezo i'm so sad

Examining Washington D.C. during the 2017 women's march and presidential inauguration, this paper looks to examine the effect that large scale public events have on urban mobility. On January 20, 2017, Donald Trump was inaugurated in Washington. The following day, January 21, hundreds of thousands of people protested Trump's inauguration through the Women's March. The women's march in Washington has been called the largest protest in the United States with an estimated 470,000 people in attendance. Crowd scientists estimate that the crowd at the Trump inauguration was about around 160,000 people, a third of the size of the protest [@wallace_parlapiano_2017]. A population increase of this size over the course of just one weekend creates a huge shock to the transportation system. This can occur through the influx of people coming into the city as well as road blocks impeding traffic. This study looks at changes in the mean travel time to the center of D.C., with the center point defined as the Washington Monument. Before modeling this impact, however, we first modeled the impact of demographic census data on travel time during normal January conditions. We then were able to observe how demographics impact mobility and how those factors are exacerbated during major shocks to the system. We found that the number of roads, distance to the Washington Monument, and the number of black residents are consistently the most significant predictors of travel time to each polygon from the Washington Monument. Further, we found that ________ ADD FINDINGS ABOUT SIGNIFICANCE DURING SHOCK!!!

@@

```{r map_acs}
dc_acs_tracts %>%
	ggplot() +
	geom_sf(mapping = aes(fill = med_income)) +
	theme_map +
	scale_fill_viridis_c()

dc_acs_tracts %>%
	ggplot() +
	geom_sf(mapping = aes(fill = black_pct)) +
	theme_map +
	scale_fill_viridis_c()
```

As shown in figure 1 and 2, it is clear that there are sharp both economic and racial divides as shown by census demographics. Whether the partitions assigned to each demographic group affect their ability to access the center or downtown area of D.C. is a key question to answer if transportation policy is focused on helping different demographics. By using a simultaneous autoregressive model on travel time to the Washington Monument during normal times, we are able to see the impact of race on D.C. area mobility. We observed that the population of black residents in a census tract is one of the few significant variables in predicting travel time to the Washington Monument. Second, major increases in population during special events can shock the city's infrastructure. For this reason, we looked at Washington because it is a city that is especially significant during major national events such as the inauguration and therefore attracts huge numbers of additional visitors. We used a similar simultaneous autoregressive model for the week of the inauguration and Women's March and then modeled the differences between the normal weeks and the inauguration and march to observe the shock on the system. Through modeling both before and during the major events, we were able to observe both demographic significance of travel time and the impact of major population changes on the system.

@@

Our goal in modeling the travel times of different census tracts to the Washington monument is twofold. First, we want to see what factors in general influence travel times; and second, we want to see if these factors determine whether a census tract is heavily affected by a signficant event such as the inauguration and Women's March. The census track features we consider during this analysis are the distance to Washington monument, median income, asian, white, and black percentage demographics, percent of people who use cars as their primary mode of transportation versus public transit, number of primary-secondary roads (see [ACS Technical Documentation, Page 3-42](https://www2.census.gov/geo/pdfs/maps-data/data/tiger/tgrshp2015/TGRSHP2015_TechDoc.pdf)), and number of transit lines. The purpose of using several features that may distinguish census tracts and specific methods is to find non-confounded estimates and accurately infer what affects travel times.

@@

As is visible in figure 1, many of the tracts neighboring each other have very similar mean travel times (MTT). The addition of transit and roadways also brings up the notion that neighboring tracts have access to similar roadways; this is especially true for the many roadways that coincide with census tract borders. We further justify the hypothesis of spatial autocorrelation through a Moran's I test. The null hypothesis of Moran's I test is that there is complete spatial randomness (CSR) which implies no spatial autocorrelation. We observe a p-value of 0 for the Moran's I statistic and thus reject the null hypothesis of CSR. In turn, we conclude the presence of spatial autocorrelation and justify the need to impose a neighborhood correlation structure in our regression models. We define two tracts to be neighbors if they simply share a portion of their border. We decide on this neighborhood structure, as opposed to within-distance and k-nearest-neighbors structures, due to the relative size of D.C's census tracts and the lack of tracts that share very minimal borders. Since tracts are relatively small, we can be confident that every bordering neighbor is physically close to any point in a tract. Also because there are not tracts that share minimal borders, we can exclude the possibility of edge cases such as where long and narrow tracts neighbor with each other at one point.

@@

@@

Our first spatial autoregressive model looks at how mean travel times is affected by the estimated demographics in each census tract before the inauguration. The coefficients `med_income`, `asian_pct`, `public_transit_pct`, `car_transit_pct`, `n_roads`, `n_transit` are all negative, implying that it takes longer to travel from the Washington Monument to the polygons in which there are lower median income, lower percentage of Asian population, and lower number of public transit, car transit, roads, and metro lines. It is usually the case that the city attributes such as `public_transit_pct`, `car_transit_pct`, `n_roads`, and `n_transit` have an immediate impact on transportation and thus can directly account for the variation in travel times, whereas the link between demographic factors in a region and travel times might not be as clear. One possible explanation is that the areas where people have lower income are often further away from the city center and less accessible by public transport, which altogether results in longer travel times and decreased mobility. The coefficients `white_pct`, `black_pct`, and `distance` are all positive, implying that it is faster to get to the closer polygons or those in which there is a lower percentage of white and black residents from the Washington Monument. Although it makes intuitive sense that distance is positively correlated with time, it does not come as natural to think about why there might be such positive relationship between racial proportion of whites and blacks with travel times. Based on the visualizations, the fact that the predominantly black areas are seen further east while the predominantly white areas are seen further west indicates that the variation observed in mean travel times can be explained by the variation in geographical distance and inherent accessibility in areas dominated by certain races. Furthermore, it is not surprising that `n_roads`, `n_transit`, and `distance` all have significant $p$-values as they are generally the primary predictors that contribute to the estimation of travel times.

>>>>

Examining Washington D.C. during the 2017 women's march and presidential inauguration, this paper looks to examine the effect that large scale public events have on urban mobility. On January 20, 2017, Donald Trump was inaugurated in Washington. The following day, January 21, hundreds of thousands of people protested Trump's inauguration through the Women's March. The women's march in Washington has been called the largest protest in the United States with an estimated 470,000 people in attendance. Crowd scientists estimate that the crowd at the Trump inauguration was about around 160,000 people, a third of the size of the protest [@wallace_parlapiano_2017]. A population increase of this size over the course of just one weekend creates a huge shock to the transportation system. This can occur through the influx of people coming into the city as well as road blocks impeding traffic. This study looks at changes in the mean travel time from the center of D.C., with the center point defined as the Washington Monument. Before modeling this impact, however, we first modeled the impact of demographic census data on travel time during normal January conditions. We then were able to observe how demographics impact mobility and how those factors are exacerbated during major shocks to the system. We found that the number of roads, distance to the Washington Monument, and the number of black residents are consistently the most significant predictors of travel time to each polygon from the Washington Monument. Further, we found that ________ ADD FINDINGS ABOUT SIGNIFICANCE DURING SHOCK!!!

@@

\begin{figure}[H]
	\centering
  \includegraphics[width=0.48\textwidth]{img/dc_income_map.png}
	\includegraphics[width=0.48\textwidth]{img/dc_black_map.png}
  \caption{(A) - Median income by census tract; (B) - Percent of Black Citizens by census tract}
\end{figure}

```{r map_acs}
dc_income_map <- dc_acs_tracts %>%
	ggplot() +
	geom_sf(mapping = aes(fill = med_income)) +
	theme_map +
	scale_fill_viridis_c() +
	theme(legend.position="bottom") + 
	guides(fill = guide_colourbar(barwidth = 20))

dc_black_map <- dc_acs_tracts %>%
	ggplot() +
	geom_sf(mapping = aes(fill = black_pct)) +
	theme_map +
	scale_fill_viridis_c() +
	theme(legend.position="bottom") + 
	guides(fill = guide_colourbar(barwidth = 20))
```

As shown in figure 1 and 2, it is clear that there are sharp both economic and racial divides as shown by census demographics. Whether the partitions assigned to each demographic group affect their ability to access the center or downtown area of D.C. is a key question to answer if transportation policy is focused on helping different demographics. By using a simultaneous autoregressive model on travel time from the Washington Monument during normal times, we are able to see the impact of race on D.C. area mobility. We observed that the population of black residents in a census tract is one of the few significant variables in predicting travel time to the Washington Monument. Second, major increases in population during special events can shock the city's infrastructure. For this reason, we looked at Washington because it is a city that is especially significant during major national events such as the inauguration and therefore attracts huge numbers of additional visitors. We used a similar simultaneous autoregressive model for the week of the inauguration and Women's March and then modeled the differences between the normal weeks and the inauguration and march to observe the shock on the system. Through modeling both before and during the major events, we were able to observe both demographic significance of travel time and the impact of major population changes on the system.

@@

Our goal in modeling the travel times of different census tracts from the Washington monument is twofold. First, we want to see what factors in general influence travel times; and second, we want to see if these factors determine whether a census tract is heavily affected by a signficant event such as the inauguration and Women's March. The census track features we consider during this analysis are the distance to Washington monument, median income, asian, white, and black percentage demographics, percent of people who use cars as their primary mode of transportation versus public transit, number of primary-secondary roads (see [ACS Technical Documentation, Page 3-42](https://www2.census.gov/geo/pdfs/maps-data/data/tiger/tgrshp2015/TGRSHP2015_TechDoc.pdf)), and number of transit lines. The purpose of using several features that may distinguish census tracts and specific methods is to find non-confounded estimates and accurately infer what affects travel times.

@@

As is visible in figure 3, many of the tracts neighboring each other have very similar mean travel times (MTT). The addition of transit and roadways also brings up the notion that neighboring tracts have access to similar roadways; this is especially true for the many roadways that coincide with census tract borders. We further justify the hypothesis of spatial autocorrelation through a Moran's I test. The null hypothesis of Moran's I test is that there is complete spatial randomness (CSR) which implies no spatial autocorrelation. We observe a p-value of 0 for the Moran's I statistic and thus reject the null hypothesis of CSR. In turn, we conclude the presence of spatial autocorrelation and justify the need to impose a neighborhood correlation structure in our regression models. We define two tracts to be neighbors if they simply share a portion of their border. We decide on this neighborhood structure, as opposed to within-distance and k-nearest-neighbors structures, due to the relative size of D.C's census tracts and the lack of tracts that share very minimal borders. Since tracts are relatively small, we can be confident that every bordering neighbor is physically close to any point in a tract. Also because there are not tracts that share minimal borders, we can exclude the possibility of edge cases such as where long and narrow tracts neighbor with each other at one point.

@@

We first look how travel times are related to the selected census track features two weeks before the inaugural weekend. The response, $\mathbf{Y}$, is the mean travel time to each census tract from the Washington Monument aggregated amongst the first two weeks of January. This response is chosen because it is representative of normal travel times in January and can thus be compared to the travel times during the inaugural weekend. The model should allow us to estimate a general travel time link to demographics, roads, and transit. Define $\mathbf{X}$ to be the matrix variables containing distance to the monument, median income, number of primary-secondary roads, number of transit lines, and demographic percentages for all census tracts. Finally, let $W$ be the binary adjacency matrix such that $w_{i,j}$, the $i^{th}$ entry in the $j^{th}$ column, is 1 if $i$ and $j$ are neighbors and 0 otherwise. The SAR model is thus:

\begin{equation}
Y = \underbrace{\lambda W Y}_{\text{Autoregressive Term}} + \underbrace{\mathbf{X}\beta}_{\text{Predictors}} + \underbrace{\epsilon}_{\text{Random Error}}, \qquad \epsilon \sim N(0, \sigma^2I)
\end{equation}

As shown in equation 1,

@@

Our first spatial autoregressive model looks at how mean travel time is affected by the estimated demographics in each census tract before the inauguration. The coefficients `med_income`, `asian_pct`, `public_transit_pct`, `car_transit_pct`, `n_roads`, `n_transit` are all negative, implying that it takes longer to travel from the Washington Monument to the polygons in which there are lower median income, lower percentage of Asian population, and lower number of public transit, car transit, roads, and metro lines. It is usually the case that the city attributes such as `public_transit_pct`, `car_transit_pct`, `n_roads`, and `n_transit` have an immediate impact on transportation and thus can directly account for the variation in travel times, whereas the link between demographic factors in a region and travel times might not be as clear. One possible explanation is that the areas where people have lower income are often further away from the city center and less accessible by public transport, which altogether results in longer travel times and decreased mobility. The coefficients `white_pct`, `black_pct`, and `distance` are all positive, implying that it is faster to get to the closer polygons or those in which there is a lower percentage of white and black residents from the Washington Monument. Although it makes intuitive sense that distance is positively correlated with time, it does not come as natural to think about why there might be such positive relationship between racial proportion of whites and blacks with travel times. Based on the visualizations, the fact that the predominantly black areas are seen further east while the predominantly white areas are seen further west indicates that the variation observed in mean travel times can be explained by the variation in geographical distance and inherent accessibility in areas dominated by certain races. Furthermore, it is not surprising that `n_roads`, `n_transit`, and `distance` all have significant $p$-values as they are generally the primary predictors that contribute to the estimation of travel times.
